---
title: "Taking the S.T.E.P. Towards a Fruitful Future: A Fiscally Grounded Macroeconomic Strategy for Britain"
subtitle: "Combining Revenue-Neutral Tax Swapping, Purpose-Driven Welfare, Regional Devolution, and GenAI Productivity to Break the Doom Loop"
date: last-modified
format:
  html:
    fig-width: 10
    fig-height: 6
    toc: false
    number-sections: false
    code-fold: true
    code-tools: true
    embed-resources: true
    css: ../../style.css
    theme: none
execute:
  echo: false
  warning: false
  message: false
  freeze: false
editor: 
  markdown: 
    wrap: 72
---

[← Back to Home](../index.html){.backlink}

# Beyond the Doom Loop: Reclaiming Economic Hope

For over a decade, Britain’s macroeconomic discourse has been trapped in
a self-fulfilling "doom loop." Stagnant Total Factor Productivity (TFP)
growth, rising structural health and disability expenditures, and an
escalating sovereign debt-service burden have created a climate of
fiscal pessimism. This pessimism feeds higher gilt yields, elevates risk
premia, suppresses private investment, and ultimately shrinks the tax
base—forcing governments into reactive fiscal tightening that further
dampens potential growth.

Breaking out of this vicious cycle requires more than incremental
adjustments; it demands a clear, novel, and self-reinforcing policy
framework. Enter the **S.T.E.P.** platform (*Switch, Tech, Empower,
Purpose*)—a package of four targeted structural reforms designed for
rapid short-term implementation (Years 1–3) without triggering uncosted
borrowing:

1.  **S – Tax Switch for Working Families:** Reverses employment tax
    squeezes by cutting Personal National Insurance by 2 percentage
    points, fully funded by equalising the tax treatment of unearned
    investment and capital income with earned income.
2.  **T – Tech & AI Growth Guarantee:** Deploys off-balance-sheet
    public-private co-investment vehicles via the National Wealth Fund
    and British Business Bank to match SME digital/AI software adoption
    and fund applied mathematics short-courses.
3.  **E – Empowering Regions (Single-Pot Devolution):** Collapses over
    30 fragmented Whitehall competitive bidding pots into single-pot
    block grants for Combined Mayoral Authorities, funded by
    rationalising central civil service overheads and active sales of
    surplus public land assets.
4.  **P – Purpose-Driven Welfare (Health-to-Work Gateway):** Replaces
    duplicate, adversarial assessment processes with a single
    Health-to-Work gateway tied directly to Integrated Care Boards,
    substituting low-level cash top-ups with direct-access vouchers and
    immediate occupational therapy triage.

### The Self-Financing Virtuous Cycle

The core innovation of the S.T.E.P. platform lies in its **inbuilt
self-financing design**. By prioritising revenue-neutral tax swaps,
Whitehall operational savings, and off-budget asset-backed investment
vehicles, the strategy protects public finances in the short term.

Crucially, as confidence returns, a positive feedback loop takes hold.
Improved business sentiment crowds in private investment, labour force
participation stabilises via health-to-work pathways, and GenAI adoption
lifts baseline productivity. The resulting expansion of the real tax
base, combined with falling sovereign borrowing yields
($r_{\text{gov}}$) and narrowing risk premia ($\pi_{\text{risk}}$),
wards off bond vigilantes and secures a sustainable, long-run fiscal
dividend.

------------------------------------------------------------------------

# The Macroeconomic Framework: Modelling the Transition

To demonstrate that this positive outcome is structurally sound rather
than a collection of random projections, we model the policy platform
using an augmented **Solow-Swan supply-side framework** coupled with
**recursive sovereign debt accumulation dynamics**.

## Supply-Side Potential Growth

Aggregate output $Y_t$ is governed by a Cobb-Douglas production function
with labour-augmenting technological progress $A_t$:

$$Y_t = K_t^\alpha (A_t L_t)^{1-\alpha}$$

Taking logarithms and time derivatives yields the standard growth
accounting identity for potential real GDP growth ($g_t$):

$$g_t = \alpha g_{K,t} + (1-\alpha)(g_{A,t} + g_{L,t})$$

where: \* $\alpha = 0.40$ represents the profit share (capital share of
aggregate income), calibrated from Penn World Table (PWT) 11.0 national
income accounts. \* $g_{A,t}$ is labour-augmenting Total Factor
Productivity (TFP) growth, which accelerates asymptotically from an
initial OBR baseline of 0.9% to a GenAI-boosted steady-state target of
**2.25%**. \* $g_{L,t}$ is labour force growth, stabilised at
**0.4%–0.5%** via PIP health-to-work reforms. \* $g_{K,t}$ is capital
stock growth, which endogenously converges toward the Solow Balanced
Growth Path (BGP) equilibrium: $g_K^* = g_A + g_L = 2.65\%$.

## Capital Returns & Public Borrowing Yields

The gross Marginal Product of Capital ($\text{MPK}$) and the net private
real return ($r_{\text{priv}}$) are derived from the aggregate
capital-output ratio ($K/Y \approx 4.8$, based on PWT 11.0 broad capital
stock data):

$$\text{MPK}_t = \alpha \left(\frac{Y_t}{K_t}\right) \approx 0.40 \times \left(\frac{1}{4.8}\right) \approx 8.3\%$$

$$r_{\text{priv},t} = \text{MPK}_t - \delta = 8.3\% - 5.0\% = 3.3\%$$

where $\delta = 5.0\%$ is the fixed rate of capital depreciation. Public
sector real borrowing yields ($r_{\text{gov},t}$) are decoupled from
private returns via a market risk/convenience wedge
($\pi_{\text{risk},t}$):

$$r_{\text{priv},t} = r_{\text{gov},t} + \pi_{\text{risk},t}$$

Anchor-initialised at $r_{\text{gov}} = 3.1\%$ in 2027 (matching the
10-year nominal gilt yield of \~5.1% minus 2.0% inflation expectations),
$r_{\text{gov}}$ falls toward 1.0% as fiscal credibility is established.

## Recursive Debt Sustainability Mechanics

The sovereign debt-to-GDP ratio ($b_t$) evolves dynamically according to
the chained law of motion:

$$b_t = (1 + r_{\text{gov},t} - g_t) b_{t-1} + pd_t$$

$$\Delta b_t = (r_{\text{gov},t} - g_t) b_{t-1} + pd_t$$

where $pd_t$ is the primary deficit ratio. Defining the **Sustainability
Gap** as $g_t - r_{\text{gov},t}$: \* **Dynamic Stability
(**$g > r_{\text{gov}}$): The economy generates a positive growth
dividend that naturally erodes the debt ratio, allowing structural
primary surpluses ($pd < 0$) to accelerate debt reduction.

------------------------------------------------------------------------

# Simulation Results & Horizon Projections

The simulation is initialised using official Office for Budget
Responsibility (OBR) baseline figures for 2025/26 ($b_{2026} = 94.0\%$,
initial primary deficit $= 1.4\%$, debt interest $= 2.9\%$, GDP growth
$= 1.5\%$). From 2028 onward, the S.T.E.P. platform structural effects
transition the UK economy toward its long-run steady state.

```{r}
library(tidyverse)
library(gt)
library(patchwork)

# ------------------------------------------------------------------------------
# 1. Base Model Calibration Parameters
# ------------------------------------------------------------------------------
alpha        <- 0.40   # Profit share / Capital share of income (PWT 11.0)
delta        <- 0.05   # Annual capital depreciation rate (5%)
pi_target    <- 0.02   # Expected inflation target (2%)
b_initial    <- 0.940  # OBR starting debt ratio at end-2026 (94.0%)
ky_init      <- 4.80   # Capital-Output Ratio (PWT 11.0 baseline)

macro_sim <- tibble(
  Period = c("2027", "2028", "2029", "2030s (Avg)", "Long Run (2040s+)"),
  tfp_growth      = c(0.90, 1.35, 1.70, 2.25, 2.25) / 100, # TFP growth
  labour_growth   = c(0.60, 0.45, 0.45, 0.50, 0.40) / 100, # Labour supply
  r_gov           = c(3.10, 2.20, 1.75, 1.25, 1.00) / 100, # Public real funding rate
  primary_deficit = c(1.40, 0.60, 0.00, -0.60, -1.20) / 100, # Primary deficit
  ky_ratio        = c(4.80, 4.80, 4.80, 4.85, 4.90)        # Capital-Output ratio
)

# ------------------------------------------------------------------------------
# 2. Dynamic Growth Accounting & Chained Debt Calculations
# ------------------------------------------------------------------------------
macro_sim <- macro_sim %>%
  mutate(eff_labour_growth = tfp_growth + labour_growth)

gK_vec <- numeric(nrow(macro_sim))
gK_vec[1] <- 0.80 * macro_sim$eff_labour_growth[1]
for(t in 2:nrow(macro_sim)) {
  adj_weight <- ifelse(t == 5, 1.0, 0.65 + (t * 0.08))
  gK_vec[t]  <- adj_weight * macro_sim$eff_labour_growth[t] + (1 - adj_weight) * gK_vec[t-1]
}

macro_sim <- macro_sim %>%
  mutate(
    capital_growth = gK_vec,
    gdp_growth = (alpha * capital_growth) + ((1 - alpha) * eff_labour_growth),
    g_minus_r = gdp_growth - r_gov,
    mpk_gross = alpha * (1 / ky_ratio),
    r_private = mpk_gross - delta,
    risk_premium = r_private - r_gov
  )

n_periods <- nrow(macro_sim)
b_start <- numeric(n_periods)
b_end   <- numeric(n_periods)
current_b <- b_initial

for (i in 1:n_periods) {
  b_start[i] <- current_b
  horizon_years <- ifelse(i == 4, 10, ifelse(i == 5, 10, 1))
  temp_b <- current_b
  for (yr in 1:horizon_years) {
    temp_b <- (1 + macro_sim$r_gov[i] - macro_sim$gdp_growth[i]) * temp_b + macro_sim$primary_deficit[i]
  }
  b_end[i] <- temp_b
  current_b <- temp_b 
}

macro_sim <- macro_sim %>%
  mutate(
    debt_start = b_start,
    debt_end = b_end,
    debt_interest_pay = r_gov * debt_start,
    headline_deficit = primary_deficit + debt_interest_pay
  )
```

```{r}

#| label: macro-dashboard
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Macroeconomic Growth Acceleration and Fiscal Sustainability Trajectory"
#| fig-alt: "Two panel chart showing Solow potential growth convergence and declining government debt dynamics."

theme_uk_policy <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 11, color = "#1a1a1a"),
      plot.subtitle = element_text(size = 8.5, color = "#4a4a4a"),
      axis.title = element_text(face = "bold", size = 9),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_line(color = "#e5e5e5"),
      panel.grid.major.y = element_line(color = "#e5e5e5"),
      plot.background = element_rect(fill = "white", color = NA)
    )
}

p1 <- ggplot(macro_sim, aes(x = Period, group = 1)) +
  geom_line(aes(y = gdp_growth * 100, color = "Real GDP Growth (g)"), linewidth = 1.2) +
  geom_point(aes(y = gdp_growth * 100, color = "Real GDP Growth (g)"), size = 2.5) +
  geom_line(aes(y = r_gov * 100, color = "Gov Real Funding Rate (r_gov)"), linewidth = 1.0, linetype = "dashed") +
  geom_point(aes(y = r_gov * 100, color = "Gov Real Funding Rate (r_gov)"), size = 2.0) +
  scale_color_manual(name = "", values = c("Real GDP Growth (g)" = "#002060", "Gov Real Funding Rate (r_gov)" = "#c0392b")) +
  labs(
    title = "A. Real Growth vs. Public Borrowing Yields",
    subtitle = "Positive sustainability gap (g > r_gov) established by 2029",
    x = NULL, y = "Percent (%)"
  ) +
  theme_uk_policy() +
  theme(legend.position = "bottom")

p2 <- ggplot(macro_sim, aes(x = Period, group = 1)) +
  geom_col(aes(y = headline_deficit * 100, fill = "Headline Deficit (% GDP)"), width = 0.4, alpha = 0.7) +
  geom_line(aes(y = debt_end * 100 / 2.5, color = "Sovereign Debt Ratio (b_t)"), linewidth = 1.2) +
  geom_point(aes(y = debt_end * 100 / 2.5, color = "Sovereign Debt Ratio (b_t)"), size = 2.5) +
  scale_y_continuous(
    name = "Headline Deficit (% GDP)",
    sec.axis = sec_axis(~ . * 2.5, name = "Sovereign Debt Ratio (% GDP)")
  ) +
  scale_fill_manual(name = "", values = c("Headline Deficit (% GDP)" = "#2980b9")) +
  scale_color_manual(name = "", values = c("Sovereign Debt Ratio (b_t)" = "#8e44ad")) +
  labs(
    title = "B. Deficit Reduction & Sovereign Debt Decay",
    subtitle = "Debt ratio unwinds cleanly from 97% to 55% of GDP",
    x = NULL
  ) +
  theme_uk_policy() +
  theme(legend.position = "bottom")

p1 | p2

```
