---
title: "Economists S.T.E.P. Forward"
subtitle: "The Cheerful Science Offers Circuit Breakers and Hope"
format:
  html:
    fig-width: 10
    toc: false
    number-sections: false
    code-fold: true
    code-tools:
      source: true
      toggle: false
      caption: none
    embed-resources: true
    css: ../style.css
    theme: ""
execute:
  echo: false
  warning: false
  message: false
---

[← Back to Home](../index.html){.backlink}



# Beyond the Doom Loop

For many years, Britain has been trapped in a self-fulfilling "doom loop." Productivity growth has stagnated. Structural health and disability expenditures have risen sharply. And, despite historically high tax rates, the fiscal backdrop just gets worse as debt interst burdens escalate. The despair feeds higher gilt yields, elevates risk premia, suppresses private investment, and ultimately shrinks the tax base. Governments get pressed into austerity mode, reinforcing the weakness in potential growth and accelerating the downward spiral.

We all know that breaking out of this vicious cycle requires more than incremental
adjustments; it demands a clear, novel, and self-reinforcing policy
framework. Enter a hypothetical **S.T.E.P.** platform (*Switch, Tech, Empower,
Purpose*) — a package of four targeted structural reforms designed for
short-term implementation (Years 1–3) as a foundation for a long-run growth revival without fiscal headaches:

1.  **S – Tax Switch for Working Families:** Reverses employment tax
    squeezes by cutting Personal National Insurance by 2 percentage
    points, funded by equalising the tax treatment of unearned
    investment and capital income with earned income.
2.  **T – Tech & AI Growth Guarantee:** Put AI, biotech and science at the heart of government and     education. Deploy co-investment (public/private, UK/European) vehicles via a National Wealth      Fund and a British Business Bank. 
3.  **E – Empowering Regions (Single-Pot Devolution):** Collapse fragmented Whitehall
    competitive bidding pots into single-pot
    block grants for Combined Mayoral Authorities, funded by
    rationalising central civil service overheads and active sales of
    surplus public land assets. Manchester (and Grimsby? self-interest declared!) to be major         West/East hubs for devolution
4.  **P – Purpose-Driven Welfare (Health-to-Work Gateway):** Replaces
    duplicate, adversarial assessment processes with a single
    Health-to-Work gateway tied directly to Integrated Care Boards,
    substituting low-level cash top-ups with direct-access vouchers and
    immediate occupational therapy triage.

Pie-in-the-sky? Maybe. But there are plenty of similarly great ideas out there. Whether the new Burnham administartion will reallly follow through remians to be seen. What is clear is that the economics of such a paradigm shift are bfar from dismal. Indeed, as the following heuristic modelling shows, the outcome could be extremely positive.

### The Self-Financing Virtuous Cycle

```{mermaid}
%%{init: {'theme': 'base', 'themeVariables': { 'primaryTextColor': '#ffffff', 'lineColor': '#5dade2', 'edgeLabelBackground':'transparent'}}}%%
flowchart LR
    subgraph STEP["THE S.T.E.P. PILLARS"]
        direction TB
        S["SWITCH<br/>(Tax Swap)"]
        T["TECH<br/>(GenAI TFP)"]
        E["EMPOWER<br/>(Mayors)"]
        P["PURPOSE<br/>(PIP Reform)"]
    end

    S --> Hope
    T --> Hope
    E --> Hope
    P --> Hope

    subgraph CYCLE[" "]
        direction TB
        Hope["Boost Sentiment<br/>& Hope"]
        Growth["Accelerate Potential<br/>Real GDP Growth (g ↑)"]
        TaxBase["Expand Tax Base<br/>& Structural Savings"]
        Vigilantes["Ward Off Bond Vigilantes<br/>r_gov ↓ risk wedge π_risk ↓"]
        Decay["Sustainable Debt Decay<br/>g > r_gov)"]

        Hope --> Growth
        Growth --> TaxBase
        TaxBase --> Vigilantes
        Vigilantes --> Decay
    end

    style STEP fill:transparent,stroke:#2980b9,stroke-width:1.5px,color:#ffffff
    style CYCLE fill:transparent,stroke:none

    style S fill:#1b4f72,stroke:#5dade2,stroke-width:1.5px,color:#ffffff
    style T fill:#1b4f72,stroke:#5dade2,stroke-width:1.5px,color:#ffffff
    style E fill:#1b4f72,stroke:#5dade2,stroke-width:1.5px,color:#ffffff
    style P fill:#1b4f72,stroke:#5dade2,stroke-width:1.5px,color:#ffffff

    style Hope fill:#1b4f72,stroke:#5dade2,stroke-width:1.5px,color:#ffffff
    style Growth fill:#1b4f72,stroke:#5dade2,stroke-width:1.5px,color:#ffffff
    style TaxBase fill:#1b4f72,stroke:#5dade2,stroke-width:1.5px,color:#ffffff
    style Vigilantes fill:#1b4f72,stroke:#5dade2,stroke-width:1.5px,color:#ffffff
    style Decay fill:#1b4f72,stroke:#5dade2,stroke-width:1.5px,color:#ffffff
```


The core attraction of a S.T.E.P. platform lies in its **inbuilt
self-financing design**. By prioritising revenue-neutral tax swaps,
Whitehall operational savings, and off-budget asset-backed investment
vehicles, the strategy protects public finances in the short term and builds a solid foundation for future sustainability.

Crucially, as confidence returns, a positive feedback loop takes hold.
Improved business sentiment crowds in private investment, labour force
participation stabilises via health-to-work pathways, and GenAI adoption
lifts baseline productivity. The resulting expansion of the real tax
base, combined with falling sovereign borrowing yields
($r_{\text{gov}}$) and narrowing risk premia ($\pi_{\text{risk}}$),
wards off bond vigilantes and secures a sustainable, long-run fiscal
dividend.

------------------------------------------------------------------------

# The Macro Modelling Framework

To demonstrate that a positive outcome is entirely feasible, we model the policy platform
in the context of standard growth and public finance models.

## Supply-Side Potential Growth

Aggregate output $Y_t$ is governed by a Cobb-Douglas production function
with labour-augmenting technological progress $A_t$:

$$Y_t = K_t^\alpha (A_t L_t)^{1-\alpha}$$

Taking logarithms and time derivatives yields the standard growth
accounting identity for potential real GDP growth ($g_t$):

$$g_t = \alpha g_{K,t} + (1-\alpha)(g_{A,t} + g_{L,t})$$

where: \* $\alpha = 0.40$ represents the profit share (capital share of
aggregate income), calibrated from Penn World Table (PWT) 11.0 national
income accounts. \* $g_{A,t}$ is labour-augmenting Total Factor
Productivity (TFP) growth, which accelerates asymptotically from an
initial OBR baseline of 0.9% to a GenAI-boosted steady-state target of
**2.25%**. \* $g_{L,t}$ is labour force growth, stabilised at
**0.4%–0.5%** via PIP health-to-work reforms. \* $g_{K,t}$ is capital
stock growth, which endogenously converges toward the Solow Balanced
Growth Path (BGP) equilibrium: $g_K^* = g_A + g_L = 2.65\%$.

## Capital Returns & Gilt Yields

The gross Marginal Product of Capital ($\text{MPK}$) and the net private
real return ($r_{\text{priv}}$) are derived from the aggregate
capital-output ratio ($K/Y \approx 4.8$, based on PWT 11.0 broad capital
stock data):

$$\text{MPK}_t = \alpha \left(\frac{Y_t}{K_t}\right) \approx 0.40 \times \left(\frac{1}{4.8}\right) \approx 8.3\%$$

$$r_{\text{priv},t} = \text{MPK}_t - \delta = 8.3\% - 5.0\% = 3.3\%$$

where $\delta = 5.0\%$ is the fixed rate of capital depreciation. Public
sector real borrowing yields ($r_{\text{gov},t}$) are decoupled from
private returns via a market risk/convenience wedge
($\pi_{\text{risk},t}$):

$$r_{\text{priv},t} = r_{\text{gov},t} + \pi_{\text{risk},t}$$

Anchor-initialised at $r_{\text{gov}} = 3.1\%$ in 2027 (matching the
10-year nominal gilt yield of \~5.1% minus 2.0% inflation expectations),
$r_{\text{gov}}$ falls toward 1.0% as fiscal credibility is established.

## Debt Sustainability Mechanics

The sovereign debt-to-GDP ratio ($b_t$) evolves dynamically:

$$b_t = (1 + r_{\text{gov},t} - g_t) b_{t-1} + pd_t$$

$$\Delta b_t = (r_{\text{gov},t} - g_t) b_{t-1} + pd_t$$

where $pd_t$ is the primary deficit ratio. 
The **Sustainability Gap** is defined as $g_t - r_{\text{gov},t}$ with a stable fiscal regime ensured once **$g > r_{\text{gov}}$

The economy then generates a positive growth
dividend that naturally erodes the debt ratio, allowing structural
primary surpluses ($pd < 0$) to accelerate debt reduction.

------------------------------------------------------------------------

# Simulation Results & Horizon Projections

The simulation is initialised using Office for Budget
Responsibility (OBR) baseline figures for 2025/26 

$b_{2026} = 94.0\%$, initial primary deficit $= 1.4\%$, debt interest $= 2.9\%$, GDP growth
$= 1.5\%$.

Into the 2030s, S.T.E.P. leads the UK toward its long-run steady state - its happy place.

```{r}
library(tidyverse)
library(gt)
library(patchwork)

# ------------------------------------------------------------------------------
# 1. Base Model Calibration Parameters
# ------------------------------------------------------------------------------
alpha        <- 0.40   # Profit share / Capital share of income (PWT 11.0)
delta        <- 0.05   # Annual capital depreciation rate (5%)
pi_target    <- 0.02   # Expected inflation target (2%)
b_initial    <- 0.940  # OBR starting debt ratio at end-2026 (94.0%)
ky_init      <- 4.80   # Capital-Output Ratio (PWT 11.0 baseline)

macro_sim <- tibble(
  Period = c("2027", "2028", "2029", "2030s (Avg)", "Long Run (2040s+)"),
  tfp_growth      = c(0.90, 1.35, 1.70, 2.25, 2.25) / 100, # TFP growth
  labour_growth   = c(0.60, 0.45, 0.45, 0.50, 0.40) / 100, # Labour supply
  r_gov           = c(3.10, 2.20, 1.75, 1.25, 1.00) / 100, # Public real funding rate
  primary_deficit = c(1.40, 0.60, 0.00, -0.60, -1.20) / 100, # Primary deficit
  ky_ratio        = c(4.80, 4.80, 4.80, 4.85, 4.90)        # Capital-Output ratio
)

# ------------------------------------------------------------------------------
# 2. Dynamic Growth Accounting & Chained Debt Calculations
# ------------------------------------------------------------------------------
macro_sim <- macro_sim %>%
  mutate(eff_labour_growth = tfp_growth + labour_growth)

gK_vec <- numeric(nrow(macro_sim))
gK_vec[1] <- 0.80 * macro_sim$eff_labour_growth[1]
for(t in 2:nrow(macro_sim)) {
  adj_weight <- ifelse(t == 5, 1.0, 0.65 + (t * 0.08))
  gK_vec[t]  <- adj_weight * macro_sim$eff_labour_growth[t] + (1 - adj_weight) * gK_vec[t-1]
}

macro_sim <- macro_sim %>%
  mutate(
    capital_growth = gK_vec,
    gdp_growth = (alpha * capital_growth) + ((1 - alpha) * eff_labour_growth),
    g_minus_r = gdp_growth - r_gov,
    mpk_gross = alpha * (1 / ky_ratio),
    r_private = mpk_gross - delta,
    risk_premium = r_private - r_gov
  )

n_periods <- nrow(macro_sim)
b_start <- numeric(n_periods)
b_end   <- numeric(n_periods)
current_b <- b_initial

for (i in 1:n_periods) {
  b_start[i] <- current_b
  horizon_years <- ifelse(i == 4, 10, ifelse(i == 5, 10, 1))
  temp_b <- current_b
  for (yr in 1:horizon_years) {
    temp_b <- (1 + macro_sim$r_gov[i] - macro_sim$gdp_growth[i]) * temp_b + macro_sim$primary_deficit[i]
  }
  b_end[i] <- temp_b
  current_b <- temp_b 
}

macro_sim <- macro_sim %>%
  mutate(
    debt_start = b_start,
    debt_end = b_end,
    debt_interest_pay = r_gov * debt_start,
    headline_deficit = primary_deficit + debt_interest_pay
  )
```

```{r}

#| label: macro-dashboard
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Macroeconomic Growth Acceleration and Fiscal Sustainability Trajectory"
#| fig-alt: "Two panel chart showing Solow potential growth convergence and declining government debt dynamics."

theme_uk_policy <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 11, color = "#1a1a1a"),
      plot.subtitle = element_text(size = 8.5, color = "#4a4a4a"),
      axis.title = element_text(face = "bold", size = 9),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_line(color = "#e5e5e5"),
      panel.grid.major.y = element_line(color = "#e5e5e5"),
      plot.background = element_rect(fill = "white", color = NA)
    )
}

p1 <- ggplot(macro_sim, aes(x = Period, group = 1)) +
  geom_line(aes(y = gdp_growth * 100, color = "Real GDP Growth (g)"), linewidth = 1.2) +
  geom_point(aes(y = gdp_growth * 100, color = "Real GDP Growth (g)"), size = 2.5) +
  geom_line(aes(y = r_gov * 100, color = "Gov Real Funding Rate (r_gov)"), linewidth = 1.0, linetype = "dashed") +
  geom_point(aes(y = r_gov * 100, color = "Gov Real Funding Rate (r_gov)"), size = 2.0) +
  scale_color_manual(name = "", values = c("Real GDP Growth (g)" = "#002060", "Gov Real Funding Rate (r_gov)" = "#c0392b")) +
  labs(
    title = "A. Real Growth vs. Public Borrowing Yields",
    subtitle = "Positive sustainability gap (g > r_gov) established by 2029",
    x = NULL, y = "Percent (%)"
  ) +
  theme_uk_policy() +
  theme(legend.position = "bottom")

p2 <- ggplot(macro_sim, aes(x = Period, group = 1)) +
  geom_col(aes(y = headline_deficit * 100, fill = "Headline Deficit (% GDP)"), width = 0.4, alpha = 0.7) +
  geom_line(aes(y = debt_end * 100 / 2.5, color = "Sovereign Debt Ratio (b_t)"), linewidth = 1.2) +
  geom_point(aes(y = debt_end * 100 / 2.5, color = "Sovereign Debt Ratio (b_t)"), size = 2.5) +
  scale_y_continuous(
    name = "Headline Deficit (% GDP)",
    sec.axis = sec_axis(~ . * 2.5, name = "Sovereign Debt Ratio (% GDP)")
  ) +
  scale_fill_manual(name = "", values = c("Headline Deficit (% GDP)" = "#2980b9")) +
  scale_color_manual(name = "", values = c("Sovereign Debt Ratio (b_t)" = "#8e44ad")) +
  labs(
    title = "B. Deficit Reduction & Sovereign Debt Decay",
    subtitle = "Debt ratio unwinds cleanly from 97% to 55% of GDP",
    x = NULL
  ) +
  theme_uk_policy() +
  theme(legend.position = "bottom")

p1 | p2
```


```{r}
#| label: macro-gt-table
#| echo: false
#| warning: false
#| message: false

# ------------------------------------------------------------------------------
# 1. Construct Transposed Data Frame for Table Output
# ------------------------------------------------------------------------------
macro_transposed <- tibble(
  Category = c(
    rep("Growth & Supply Side %yoy", 4),
    rep("Real Interest Rates %", 3),
    rep("Government Finances % of GDP", 5)
  ),
  Variable = c(
    "Real GDP", "Labour Force", "Capital Stock", "Total Factor Productivity",
    "Private Sector Real Return", "Risk Premium", "Government Real Funding Rate",
    "Primary Deficit", "Debt Interest Costs", "Headline Deficit", "Debt Ratio, end-period", "Sustainability Gap*"
  ),
  `2027` = c(
    macro_sim$gdp_growth[1], macro_sim$labour_growth[1], macro_sim$capital_growth[1], macro_sim$tfp_growth[1],
    macro_sim$r_private[1], macro_sim$risk_premium[1], macro_sim$r_gov[1],
    macro_sim$primary_deficit[1], macro_sim$debt_interest_pay[1], macro_sim$headline_deficit[1], macro_sim$debt_end[1], macro_sim$g_minus_r[1]
  ) * 100,
  `2028` = c(
    macro_sim$gdp_growth[2], macro_sim$labour_growth[2], macro_sim$capital_growth[2], macro_sim$tfp_growth[2],
    macro_sim$r_private[2], macro_sim$risk_premium[2], macro_sim$r_gov[2],
    macro_sim$primary_deficit[2], macro_sim$debt_interest_pay[2], macro_sim$headline_deficit[2], macro_sim$debt_end[2], macro_sim$g_minus_r[2]
  ) * 100,
  `2029` = c(
    macro_sim$gdp_growth[3], macro_sim$labour_growth[3], macro_sim$capital_growth[3], macro_sim$tfp_growth[3],
    macro_sim$r_private[3], macro_sim$risk_premium[3], macro_sim$r_gov[3],
    macro_sim$primary_deficit[3], macro_sim$debt_interest_pay[3], macro_sim$headline_deficit[3], macro_sim$debt_end[3], macro_sim$g_minus_r[3]
  ) * 100,
  `2030s (Avg)` = c(
    macro_sim$gdp_growth[4], macro_sim$labour_growth[4], macro_sim$capital_growth[4], macro_sim$tfp_growth[4],
    macro_sim$r_private[4], macro_sim$risk_premium[4], macro_sim$r_gov[4],
    macro_sim$primary_deficit[4], macro_sim$debt_interest_pay[4], macro_sim$headline_deficit[4], macro_sim$debt_end[4], macro_sim$g_minus_r[4]
  ) * 100,
  `Long Run (2040s+)` = c(
    macro_sim$gdp_growth[5], macro_sim$labour_growth[5], macro_sim$capital_growth[5], macro_sim$tfp_growth[5],
    macro_sim$r_private[5], macro_sim$risk_premium[5], macro_sim$r_gov[5],
    macro_sim$primary_deficit[5], macro_sim$debt_interest_pay[5], macro_sim$headline_deficit[5], macro_sim$debt_end[5], macro_sim$g_minus_r[5]
  ) * 100
)

# ------------------------------------------------------------------------------
# 2. Construct Dynamic Footnote Text (With LaTeX Escaped % Symbols)
# ------------------------------------------------------------------------------
dynamic_source_note <- sprintf(
  "**Notes & Model Specifications:**<br>
   1. **OBR Alignment & Production Function:** Initialised using official OBR baseline data ($b_{2026} = %.1f\\%%$, primary deficit $= %.1f\\%%$, debt interest $= %.1f\\%%$, GDP growth $= %.1f\\%%$). <br>Production technology assumes Cobb-Douglas with profit share $\\alpha = %.0f\\%%$, capital depreciation $\\delta = %.1f\\%%$, and Penn World Table 11.0 $K/Y \\approx %.1f$. Gross $\\text{MPK} = \\alpha (Y/K) = %.1f\\%%$.<br>
   2. **Debt & Inflation:** Inflation expectations anchored at $\\pi = %.1f\\%%$; corresponding nominal 10-year funding rate is $i = r_{\\text{gov}} + \\pi = %.1f\\%%$ in 2027.<br>
   3. ***Sustainability Gap Definition:** Real GDP Growth less Government Real Funding Rate ($g - r_{\\text{gov}}$). Positive indicates dynamic stability ($g > r_{\\text{gov}}$).<br>
   4. **Solow Balanced Growth Path (BGP):** Final column represents BGP steady state ($g_K^* = g_A + g_L = %.2f\\%%$, yielding $g^* = %.2f\\%%$).",
  b_initial * 100,                                    # 1. b_2026 (%)
  macro_sim$primary_deficit[1] * 100,                 # 2. Primary Deficit (%)
  macro_sim$r_gov[1] * b_initial * 100,               # 3. Debt Interest (%)
  macro_sim$gdp_growth[1] * 100,                      # 4. GDP Growth (%)
  alpha * 100,                                        # 5. Alpha Profit Share (%)
  delta * 100,                                        # 6. Depreciation Delta (%)
  ky_init,                                            # 7. K/Y Ratio (No %)
  macro_sim$mpk_gross[1] * 100,                       # 8. Gross MPK (%)
  pi_target * 100,                                    # 9. Inflation Target pi (%)
  (macro_sim$r_gov[1] + pi_target) * 100,             # 10. Nominal Funding Rate i (%)
  (macro_sim$tfp_growth[5] + macro_sim$labour_growth[5]) * 100, # 11. BGP Effective Labour (%)
  (macro_sim$gdp_growth[5]) * 100                     # 12. BGP Steady-State g* (%)
)

# ------------------------------------------------------------------------------
# 3. Render Table directly to Quarto HTML Pipeline
# ------------------------------------------------------------------------------
gt_a4_table <- macro_transposed %>%
  gt(groupname_col = "Category") %>%
  tab_header(
    title = md("**Solow-BGP Macroeconomic & Dynamic Fiscal Simulation**"),
    subtitle = md(sprintf("Initialised to OBR Baseline ($b_{2026} = %.1f%%$, Profit Share $\\alpha = %.0f%%$)", b_initial * 100, alpha * 100))
  ) %>%
  cols_label(Variable = "Macroeconomic Indicator") %>%
  
  # --- Column Width Fine-Tuning ---
  # Adjust px or % values below to balance layout across desktop & mobile
  cols_width(
    Variable ~ px(260),
    `2027` ~ px(85),
    `2028` ~ px(85),
    `2029` ~ px(85),
    `2030s (Avg)` ~ px(110),
    `Long Run (2040s+)` ~ px(140)
  ) %>%
  
  # --- Number Formatting (With % suffix) ---
  fmt_number(
    columns = -c(Category, Variable), 
    decimals = 1,
    pattern = "{x}%"  # Appends % directly to table cell values (e.g. 1.4%)
  ) %>%
  
  # --- Header & Group Styles ---
  tab_style(
    style = cell_text(weight = "bold"), 
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#111827"), cell_text(weight = "bold", color = "#60a5fa")),
    locations = cells_row_groups(groups = "Growth & Supply Side %yoy")
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#111827"), cell_text(weight = "bold", color = "#f59e0b")),
    locations = cells_row_groups(groups = "Real Interest Rates %")
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#111827"), cell_text(weight = "bold", color = "#34d399")),
    locations = cells_row_groups(groups = "Government Finances % of GDP")
  ) %>%
  tab_source_note(source_note = md(dynamic_source_note)) %>%
  tab_style(
    style = cell_text(
    size = px(8),
    color = "#111827"),
    locations = cells_source_notes()) %>%
  opt_css(css = ".gt_sourcenote {line-height: 1.1;}")

# Call the object directly for Quarto execution
gt_a4_table
```
