---
title: "Economists S.T.E.P. Forward"
subtitle: "The Cheerful Science Offers Circuit Breakers and Hope"
format:
  html:
    fig-width: 10
    toc: false
    number-sections: false
    code-fold: true
    code-tools:
      source: true
      toggle: false
      caption: none
    embed-resources: true
    css: ../style.css
    theme: ""
execute:
  echo: false
  warning: false
  message: false
---

[← Back to Home](../index.html){.backlink}



# Beyond the Doom Loop

For many years, Britain has been trapped in a self-fulfilling "doom loop." Productivity growth has stagnated. Structural health and disability expenditures have risen sharply. And, despite historically high tax rates, the fiscal backdrop just gets worse. Debt interest burdens have escalated, and there are accelerating pressures on the public purse from climate change and worsening geopolitical tensions. The despair feeds higher gilt yields, suppresses private spending and growth, and ultimately shrinks the tax base. Governments then feel forced to shrink into austerity mode, reinforcing the weakness in potential growth and accelerating the downward spiral.

We all know that breaking out of this vicious cycle requires more than incremental adjustments; it demands radical, innovative change.

Who knows whether the Burnham administration will, or can, deliver the required policy changes within a timeframe, and with consequences, that are acceptable to an impatient, irascible electorate. Initial signs are encouraging but there is no guarantee of success. 

What follows is a humble economist's attempt to identify a structural reform agenda that addresses the macro changes that need to happen. It does not claim originality of policy proposals but, in exploring feasible outcomes through macro-finance models, reveals that we can dare to dream. 

## S.T.E.P. Forward

Enter a hypothetical **S.T.E.P.** platform (*Switch, Tech, Empower,Purpose*) - a package of four targeted structural reforms designed for
short-term implementation (Years 1-3) as a foundation for a long-run growth revival without fiscal headaches:

1.  **S = Switch and substitute revenue sources:** Rebalance the tax burden from income to wealth. This includes cutting employee National Insurance contributions, funded by equalising the tax treatment of unearned investment and capital income with earned income. Land taxes will also play a role in re-incentivising work and thus labour force participation. Focus on immovables, anticipate and address behavioural avoidance.
2.  **T = Tech focus:** Put AI, robotics, digitalisation, biotech and science at the heart of government, education and worker skillsets. Deploy co-investment (public/private, UK/European) vehicles via a National Wealth Fund and a British Business Bank. Ensure that such vehicles enable the scaling of superstars and are not simply shelters for unproductive zombie companies.Take advantage of homegrown expertise in AI, biotech, quantum computing and financial services to maximise employment opportunities, and above all, significantly raise productivity growth. Promoting innovation clusters and accelerating diffusion of new ideas will be key. 
3.  **E = Empowering Regions:** Fully activate the revived National Economic Council. Empower mayors as economic actors, not simply Town Hall administrators. Joined-up thinking and action across regions, coupled with enhanced democratic legitimacy through devolution, will raise productivity and boost both business and consumer confidence.Tackle NHS/Social Care overlaps efficiently at regional level. 
4.  **P = Purposeful participation:** Engage fully in welfare reform. Replace duplicate, adversarial assessment processes with simplified Health-to-Work gateways. Substitute low-level cash top-ups with direct-access vouchers and immediate occupational therapy. Proactively support NEETS at regional and national levels. 

## The Self-Financing Virtuous Cycle

```{mermaid}
%%{init: {
  'theme': 'base',
  'themeVariables': {
    'fontSize': '32px',
    'fontFamily': 'Arial, sans-serif',
    'primaryColor': '#dceef8',
    'primaryTextColor': '#102a43',
    'primaryBorderColor': '#2980b9',
    'lineColor': '#2980b9',
    'edgeLabelBackground': '#ffffff',
    'clusterBkg': 'transparent',
    'clusterBorder': '#2980b9'
  },
  'flowchart': {
    'htmlLabels': true,
    'nodeSpacing': 30,
    'rankSpacing': 40
  }
}}%%

flowchart LR
    subgraph STEP["THE S.T.E.P. PILLARS"]
        direction TB
        S["SWITCH<br/>Tax Rebalancing"]
        T["TECH<br/>TFP GenAI"]
        E["EMPOWER<br/>Devolution"]
        P["PURPOSE<br/>Welfare Reform"]
    end

    S --> Hope
    T --> Hope
    E --> Hope
    P --> Hope

    subgraph CYCLE[" "]
        direction TB
        Hope["Boost Sentiment<br/>& Hope"]
        Growth["Accelerate Potential<br/>Real GDP Growth"]
        TaxBase["Expand Tax Base<br/>& Structural Savings"]
        Vigilantes["Ward Off Bond Vigilantes<br/>Real Gilt Yields fall"]
        Decay["Sustainable Debt Decay<br/>GDP Growth exceeds Financing Rate"]

        Hope --> Growth
        Growth --> TaxBase
        TaxBase --> Vigilantes
        Vigilantes --> Decay
    end

    classDef processBox fill:#dceef8,stroke:#2980b9,stroke-width:2px,color:#102a43;
    class S,T,E,P,Hope,Growth,TaxBase,Vigilantes,Decay processBox;

    style STEP fill:transparent,stroke:#2980b9,stroke-width:2px,color:#102a43
    style CYCLE fill:transparent,stroke:none
```


The core attraction of this S.T.E.P. platform lies in its **inbuilt
self-financing design** together with its capacity to break out of the doom loop and enter a reinforcing virtuous cycle. 

Crucially, as confidence returns, a positive feedback loop takes hold.
Improved sentiment crowds in private spending, labour force
participation stabilises via health-to-work pathways, and GenAI adoption
lifts baseline productivity. The strategy protects public finances in the short term and builds a solid foundation for future sustainability.

The resulting expansion of the real tax
base, combined with falling sovereign borrowing yields
and narrowing risk premia, wards off bond vigilantes and secures a sustainable, long-run fiscal
dividend.

------------------------------------------------------------------------

# The Macro Modelling Framework

To demonstrate that a positive outcome is feasible, we model the policy platform
in the context of standard growth and public finance models.

## Supply-Side Potential Growth

Aggregate output $Y_t$ is governed by a Cobb-Douglas production function
with labour-augmenting technological progress $A_t$:

$$Y_t = K_t^\alpha (A_t L_t)^{1-\alpha}$$

Taking logarithms and time derivatives yields the standard growth
accounting identity for potential real GDP growth ($g_t$):

$$g_t = \alpha g_{K,t} + (1-\alpha)(g_{A,t} + g_{L,t})$$

where:<br>
$\alpha = 0.40$ represents the profit share (capital share of
aggregate income), calibrated from Penn World Table (PWT) 11.0 estimates<br><br>
$g_{A,t}$ is labour-augmenting Total Factor
Productivity (TFP) growth, which accelerates asymptotically from an
initial OBR baseline of 0.9% to a GenAI-boosted steady-state target of
**2.25%**<br><br>
$g_{L,t}$ is labour force growth, stabilised at
**0.4%–0.5%** via PIP health-to-work and NEETs reforms<br><br>
$g_{K,t}$ is capital stock growth, which endogenously converges toward the Solow Balanced
Growth Path (BGP) equilibrium: $g_K^* = g_A + g_L = 2.6\%$.

## Capital Returns & Gilt Yields

The gross Marginal Product of Capital ($\text{MPK}$) and the net private
real return ($r_{\text{priv}}$) are derived from the aggregate
capital-output ratio $K/Y$ \approx 4.8, based on PWT 11.0 estimates for the UK:

$$\text{MPK}_t = \alpha \left(\frac{Y_t}{K_t}\right) \approx 0.40 \times \left(\frac{1}{4.8}\right) \approx 8.3\%$$

$$r_{\text{priv},t} = \text{MPK}_t - \delta = 8.3\% - 5.0\% = 3.3\%$$

where $\delta = 5.0\%$ is the assumed fixed rate of capital depreciation.

Public sector real borrowing yields ($r_{\text{gov},t}$) differ from
private returns due to a market risk/convenience wedge ($\pi_{\text{risk},t}$):

$$r_{\text{priv},t} = r_{\text{gov},t} + \pi_{\text{risk},t}$$

Initial conditions for $r_{\text{gov}} = 3.1\%$ use the current 10-year nominal gilt yield of \~5.1% less assumed inflation expectations of 2.0%.

$r_{\text{gov}}$ falls toward 1.0% as fiscal credibility is established with both domestic and international financial markets.

## Debt Sustainability Mechanics

The sovereign debt-to-GDP ratio ($b_t$) evolves dynamically:

$$b_t = (1 + r_{\text{gov},t} - g_t) b_{t-1} + pd_t$$

$$\Delta b_t = (r_{\text{gov},t} - g_t) b_{t-1} + pd_t$$

where $pd_t$ is the primary deficit ratio. 
The **Sustainability Gap** is defined as $g_t - r_{\text{gov},t}$.
Currently, Britain is in an unstable regime, signified by a negative gap. But, as the S.T.E.P program takes effect, the gap reverses into positive territory $g > r_{\text{gov}}.$ The economy then generates a positive growth dividend that naturally erodes the debt ratio, allowing structural primary surpluses ($pd < 0$) to accelerate debt reduction.

------------------------------------------------------------------------

## Simulation Results & Horizon Projections

The following simulation is calibrated using [Office for Budget Responsibility](https://obr.uk/frs/fiscal-risks-and-sustainability-july-2026/) (OBR) baseline figures for 2025/26 and estimates made in [Penn World Tables 11.0.](https://www.rug.nl/ggdc/productivity/pwt/?lang=en)  

Initial conditions are as follows: $b_{2026} = 94.0\%$ of GDP, initial primary deficit $= 1.4\%$ of GDP, debt interest payments $= 2.9\%$ of GDP, GDP growth  $= 1.5\%$yoy.

As expected, a successful and sustained S.T.E.P. implementation leads the UK toward a desirable long-run steady state, <b>a happy place</b>.

```{r}
library(tidyverse)
library(gt)
library(patchwork)

# ------------------------------------------------------------------------------
# 1. Base Model Calibration Parameters
# ------------------------------------------------------------------------------
alpha        <- 0.40   # Profit share / Capital share of income (PWT 11.0)
delta        <- 0.05   # Annual capital depreciation rate (5%)
pi_target    <- 0.02   # Expected inflation target (2%)
b_initial    <- 0.940  # OBR starting debt ratio at end-2026 (94.0%)
ky_init      <- 4.80   # Capital-Output Ratio (PWT 11.0 baseline)

macro_sim <- tibble(
  Period = c("2027", "2028", "2029", "2030s (Avg)", "Long Run (2040s+)"),
  tfp_growth      = c(0.90, 1.35, 1.70, 2.25, 2.25) / 100, # TFP growth
  labour_growth   = c(0.60, 0.45, 0.45, 0.50, 0.40) / 100, # Labour supply
  r_gov           = c(3.10, 2.20, 1.75, 1.25, 1.00) / 100, # Public real funding rate
  primary_deficit = c(1.40, 0.60, 0.00, -0.60, -1.20) / 100, # Primary deficit
  ky_ratio        = c(4.80, 4.80, 4.80, 4.85, 4.90)        # Capital-Output ratio
)

# ------------------------------------------------------------------------------
# 2. Dynamic Growth Accounting & Chained Debt Calculations
# ------------------------------------------------------------------------------
macro_sim <- macro_sim %>%
  mutate(eff_labour_growth = tfp_growth + labour_growth)

gK_vec <- numeric(nrow(macro_sim))
gK_vec[1] <- 0.80 * macro_sim$eff_labour_growth[1]
for(t in 2:nrow(macro_sim)) {
  adj_weight <- ifelse(t == 5, 1.0, 0.65 + (t * 0.08))
  gK_vec[t]  <- adj_weight * macro_sim$eff_labour_growth[t] + (1 - adj_weight) * gK_vec[t-1]
}

macro_sim <- macro_sim %>%
  mutate(
    capital_growth = gK_vec,
    gdp_growth = (alpha * capital_growth) + ((1 - alpha) * eff_labour_growth),
    g_minus_r = gdp_growth - r_gov,
    mpk_gross = alpha * (1 / ky_ratio),
    r_private = mpk_gross - delta,
    risk_premium = r_private - r_gov
  )

n_periods <- nrow(macro_sim)
b_start <- numeric(n_periods)
b_end   <- numeric(n_periods)
current_b <- b_initial

for (i in 1:n_periods) {
  b_start[i] <- current_b
  horizon_years <- ifelse(i == 4, 10, ifelse(i == 5, 10, 1))
  temp_b <- current_b
  for (yr in 1:horizon_years) {
    temp_b <- (1 + macro_sim$r_gov[i] - macro_sim$gdp_growth[i]) * temp_b + macro_sim$primary_deficit[i]
  }
  b_end[i] <- temp_b
  current_b <- temp_b 
}

macro_sim <- macro_sim %>%
  mutate(
    debt_start = b_start,
    debt_end = b_end,
    debt_interest_pay = r_gov * debt_start,
    headline_deficit = primary_deficit + debt_interest_pay
  )
```

```{r}
#| label: macro-dashboard
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Macroeconomic Growth Acceleration and Fiscal Sustainability Trajectory"
#| fig-alt: "Two panel chart showing Solow potential growth convergence and declining government debt dynamics."

library(tidyverse)
library(gt)
library(patchwork)

theme_uk_policy <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 14, color = "#1a1a1a"),
      plot.subtitle = element_text(size = 12, color = "#4a4a4a"),
      axis.title = element_text(face = "bold", size = 12),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_blank(),
      # panel.grid.major.x = element_line(color = "#e5e5e5"),
      # panel.grid.major.y = element_line(color = "#e5e5e5"),
      plot.background = element_rect(fill = "white", color = NA)
    )
}

p1 <- ggplot(macro_sim, aes(x = Period, group = 1)) +
  geom_line(aes(y = gdp_growth * 100, color = "Real GDP Growth (g)"), linewidth = 2.0) +
  geom_point(aes(y = gdp_growth * 100, color = "Real GDP Growth (g)"), size = 5) +
  geom_line(aes(y = r_gov * 100, color = "Gov Real Funding Rate (r_gov)"), linewidth = 2.0, linetype = "dashed") +
  geom_point(aes(y = r_gov * 100, color = "Gov Real Funding Rate (r_gov)"), size = 5) +
  scale_color_manual(name = "", values = c("Real GDP Growth (g)" = "#002060", "Gov Real Funding Rate (r_gov)" = "#c0392b")) +
  labs(
    title = "A. Real Growth vs. Public Borrowing Yields",
    subtitle = "Positive sustainability gap (g > r_gov) established by 2029",
    x = NULL, y = "Percent (%)"
  ) +
  theme_uk_policy() +
  theme(legend.position = "bottom")

p2 <- ggplot(macro_sim, aes(x = Period, group = 1)) +
  geom_col(
    aes(
      y = debt_end * 100,
      fill = "Government Debt Ratio (% GDP)"
    ),
    width = 0.8,
    alpha = 0.9
  ) +
  scale_y_continuous(
    name = "% of GDP"
  ) +
  scale_fill_manual(
    name = NULL,
    values = c(
      "Government Debt Ratio (% GDP)" = "#8e44ad"
    )
  ) +
  labs(
    title = "B. Sovereign Debt Decay",
    subtitle = "Debt ratio unwinds cleanly from 97% to 55% of GDP",
    x = NULL
  ) +
  theme_uk_policy() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )

p1 | p2
```


```{r}
#| label: macro-gt-table
#| echo: false
#| warning: false
#| message: false

# ------------------------------------------------------------------------------
# 1. Construct Transposed Data Frame for Table Output
# ------------------------------------------------------------------------------
macro_transposed <- tibble(
  Category = c(
    rep("Growth & Supply Side %yoy", 4),
    rep("Real Interest Rates %", 3),
    rep("Government Finances % of GDP", 5)
  ),
  Variable = c(
    "Real GDP", "Labour Force", "Capital Stock", "Total Factor Productivity",
    "Private Sector Net Real Return", "Risk Premium", "Government Real Funding Rate",
    "Primary Deficit", "Debt Interest Costs", "Headline Deficit", "Debt Ratio, end-period", "Sustainability Gap"
  ),
  `2027` = c(
    macro_sim$gdp_growth[1], macro_sim$labour_growth[1], macro_sim$capital_growth[1], macro_sim$tfp_growth[1],
    macro_sim$r_private[1], macro_sim$risk_premium[1], macro_sim$r_gov[1],
    macro_sim$primary_deficit[1], macro_sim$debt_interest_pay[1], macro_sim$headline_deficit[1], macro_sim$debt_end[1], macro_sim$g_minus_r[1]
  ) * 100,
  `2028` = c(
    macro_sim$gdp_growth[2], macro_sim$labour_growth[2], macro_sim$capital_growth[2], macro_sim$tfp_growth[2],
    macro_sim$r_private[2], macro_sim$risk_premium[2], macro_sim$r_gov[2],
    macro_sim$primary_deficit[2], macro_sim$debt_interest_pay[2], macro_sim$headline_deficit[2], macro_sim$debt_end[2], macro_sim$g_minus_r[2]
  ) * 100,
  `2029` = c(
    macro_sim$gdp_growth[3], macro_sim$labour_growth[3], macro_sim$capital_growth[3], macro_sim$tfp_growth[3],
    macro_sim$r_private[3], macro_sim$risk_premium[3], macro_sim$r_gov[3],
    macro_sim$primary_deficit[3], macro_sim$debt_interest_pay[3], macro_sim$headline_deficit[3], macro_sim$debt_end[3], macro_sim$g_minus_r[3]
  ) * 100,
  `2030s (Avg)` = c(
    macro_sim$gdp_growth[4], macro_sim$labour_growth[4], macro_sim$capital_growth[4], macro_sim$tfp_growth[4],
    macro_sim$r_private[4], macro_sim$risk_premium[4], macro_sim$r_gov[4],
    macro_sim$primary_deficit[4], macro_sim$debt_interest_pay[4], macro_sim$headline_deficit[4], macro_sim$debt_end[4], macro_sim$g_minus_r[4]
  ) * 100,
  `Long Run (2040s+)` = c(
    macro_sim$gdp_growth[5], macro_sim$labour_growth[5], macro_sim$capital_growth[5], macro_sim$tfp_growth[5],
    macro_sim$r_private[5], macro_sim$risk_premium[5], macro_sim$r_gov[5],
    macro_sim$primary_deficit[5], macro_sim$debt_interest_pay[5], macro_sim$headline_deficit[5], macro_sim$debt_end[5], macro_sim$g_minus_r[5]
  ) * 100
)

# ------------------------------------------------------------------------------
# 2. Construct Dynamic Footnote Text (With LaTeX Escaped % Symbols)
# ------------------------------------------------------------------------------
dynamic_source_note <- sprintf(
  "**Notes & Model Specifications:**<br><br>
   1. **Macro Supply:** Production technology assumes a Cobb-Douglas format with profit share $\\alpha = %.0f\\%%$, capital depreciation $\\delta = %.1f\\%%$, and $K/Y \\approx %.1f$.<br>Gross $\\text{MPK} = \\alpha (Y/K) = %.1f\\%%$<br><br>
   2. **Debt & Inflation:** Inflation expectations anchored at $\\pi = %.1f\\%%$; corresponding nominal 10-year funding rate is $i = r_{\\text{gov}} + \\pi = %.1f\\%%$ in 2027<br><br>
   3. **Sustainability Gap:** Real GDP Growth less Government Real Funding Rate ($g - r_{\\text{gov}}$).<br>Positive values ($g > r_{\\text{gov}}$) indicate dynamic stability, negative values imply instability<br><br>
   4. **Solow Balanced Growth Path:** Final column represents the model's steady state<br> $g_K^* = g_A + g_L = %.1f\\%%$, yielding $g^* = %.1f\\%%$",

  alpha * 100,                                        
  delta * 100,                                        
  ky_init,                                            
  macro_sim$mpk_gross[1] * 100,                       
  pi_target * 100,                                    
  (macro_sim$r_gov[1] + pi_target) * 100,             
  (macro_sim$tfp_growth[5] +
     macro_sim$labour_growth[5]) * 100,               
  macro_sim$gdp_growth[5] * 100                       
)

# ------------------------------------------------------------------------------
# 3. Render Table directly to Quarto HTML Pipeline
# ------------------------------------------------------------------------------
gt_a4_table <- macro_transposed %>%
  gt(groupname_col = "Category") %>%
  tab_header(
    title = md("**Illustrative Macro-Fiscal Simulation: Benchmarks NOT Forecasts**"),
    subtitle = md("Initialised using OBR and Penn World Tables 11.0 estimates")) %>%
  cols_label(Variable = "") %>%
  
  # --- Column Width Fine-Tuning ---
  # Adjust px or % values below to balance layout across desktop & mobile
  cols_width(
    Variable ~ px(260),
    `2027` ~ px(85),
    `2028` ~ px(85),
    `2029` ~ px(85),
    `2030s (Avg)` ~ px(110),
    `Long Run (2040s+)` ~ px(140)
  ) %>%
  
  # --- Number Formatting (With % suffix) ---
  fmt_number(
    columns = -c(Category, Variable), 
    decimals = 1,
    pattern = "{x}%"  # Appends % directly to table cell values (e.g. 1.4%)
  ) %>%
  
  # --- Header & Group Styles ---
  tab_style(
    style = cell_text(weight = "bold"), 
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#111827"), cell_text(weight = "bold", color = "#60a5fa")),
    locations = cells_row_groups(groups = "Growth & Supply Side %yoy")
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#111827"), cell_text(weight = "bold", color = "#f59e0b")),
    locations = cells_row_groups(groups = "Real Interest Rates %")
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#111827"), cell_text(weight = "bold", color = "#34d399")),
    locations = cells_row_groups(groups = "Government Finances % of GDP")
  ) %>%
  tab_source_note(source_note = md(dynamic_source_note)) %>%
  tab_style(
    style = cell_text(
    size = px(16),
    color = "#111827"),
    locations = cells_source_notes()) %>%
  opt_css(css = ".gt_sourcenote {line-height: 1.2;}")

# Call the object directly for Quarto execution
gt_a4_table
```

# Take That
Britain could limp along towards the next election. But who wants that, and the scary political dangers posed for socio-economic wellbeing? There is clearly appetite for radical change and the S.T.E.P framework offers workable ideas. For those worried about whether it all adds up, the macro-finance simulations show that it can - in terms of combining strong growth, better living standards, stable inflation and sustainable public finances. 

Boosting productivity holds the key to exiting the Doom Loop. Easy to say but notoriously harder to do. AI and similarly progressive technologies offer a lifeline. But while tech creation is essential, adoption and diffusion are necessary. In this regard, rebalancing taxes, meaningful devolution, offering purpose and participation for all potential workers, thoroughly reforming the health and social system jointly unlock the door to the Virtuous Cycle, leading to the happy place that we all crave.   

::: {.bibliography}

<p class="bibliography-title">Further Reading</p>

Resolution Foundation (2025) [*Call of duties*](https://www.resolutionfoundation.org/publications/call-of-duties)

Resolution Foundation (2026) [*See it. Say it. Sort it: How to reset Britain’s economic policy*](https://www.resolutionfoundation.org/app/uploads/2026/05/See-it.-Say-it.-Sort-it.pdf)

Resolution Foundation (2026) [*Mountain Climbing*](https://www.resolutionfoundation.org/app/uploads/2026/01/Mountain-climbing.pdf)
   
Resolution Foundation (2026) [*Lost in Transition*](https://www.resolutionfoundation.org/app/uploads/2026/04/Lost-in-transition.pdf)
   
IFS (2025) [*How to fix property taxes*](https://ifs.org.uk/articles/how-fix-property-taxes)

IPPR (2026) [*Turning energy support into investment leverage*](https://www.ippr.org/articles/turning-energy-support-into-investment-leverage)

Joseph Rowntree Foundation (2026) [*Tackle economic failures at root to ease pressure on Universal Credit*](https://www.jrf.org.uk/social-security/tackle-economic-failures-at-root-to-ease-pressure-on-universal-credit)
   
Centre for Cities (2026) [*Fiscal Devolution for England*](https://www.centreforcities.org/devolution/)
   
New Economics Foundation (2026) [*Reclaiming Our Regional Economies*](https://neweconomics.org/campaigns/reclaiming-our-regional-economies)
   
Centre for Policy Studies (2026) [*Patently Absurd: Why Britain is No Longer an Innovation Nation*](https://cps.org.uk/wp-content/uploads/2026/04/CPS_PATENTLY_ABSURD-final.pdf)
   
Centre for Policy Studies (2026) [*Why Raising Capital Taxes Won't Work*](https://cps.org.uk/wp-content/uploads/2026/07/Why-Raising-Capital-Gains-Tax-Wont-Work.pdf)
   
Centre for Policy Studies (2026) [*PA Productivity Toolkit for Britain*](https://cps.org.uk/wp-content/uploads/2026/02/CPS_A5_A_PRODUCTIVITY_TOOLKIT_FOR_BRITAIN.pdf)

UK Government (2026) [*No10 North/revived National Economic Council*](https://www.gov.uk/government/news/no10-north-open-for-business-as-the-governments-new-situation-room-for-making-britain-better)

UK Government (2026 update) [*The Timms Review: Interim*](https://www.gov.uk/government/collections/the-timms-review)

UK Government (2026 update) [*The Milburn Report: Interim*](https://www.gov.uk/government/publications/young-people-and-work-interim-report/young-people-and-work-interim-report)

Commission on Funding of Care and Support (2011) [*Dilnot Report*](https://www.kingsfund.org.uk/insight-and-analysis/briefings/dilnot-commission-report-social-care?) 

King's Fund (2014) [Barker Commission: A New Settlement for Health and Social Care](https://www.kingsfund.org.uk/insight-and-analysis/reports/new-settlement-health-social-care-final-report)

Care Quality Commission (2025) [State of Care 2024/25](https://www.cqc.org.uk/sites/default/files/2025-10/20251024_stateofcare2425_print.pdf)

HC Health and Social Care Committee (2025) [Adult Social Care Reform: The Cost of Inaction](https://committees.parliament.uk/publications/47713/documents/249329/default/)

Casey Commission (2025-2028) [Adult Social Care](https://caseycommission.co.uk/)

[Penn World Tables 11.0](https://www.rug.nl/ggdc/productivity/pwt/?lang=en) 

[OBR Fiscal Risks & Sustainability (Jul 2026)](https://www.gov.uk/government/publications/fiscal-risks-and-sustainability-july-2026) 
:::
