---
title: "Economists S.T.E.P. Forward"
subtitle: "The Cheerful Science Offers Circuit Breakers and Hope"
format:
  html:
    fig-width: 10
    toc: false
    number-sections: false
    code-fold: true
    code-tools:
      source: true
      toggle: false
      caption: none
    embed-resources: true
    css: ../style.css
    theme: ""
execute:
  echo: false
  warning: false
  message: false
---

[← Back to Home](../index.html){.backlink}



# Beyond the Doom Loop

For many years, Britain has been trapped in a self-fulfilling "doom loop." Productivity growth has stagnated. Structural health and disability expenditures have risen sharply. And, despite historically high tax rates, the fiscal backdrop just gets worse as debt interest burdens escalate. The despair feeds higher gilt yields, elevates risk premia, suppresses private spending and growth, and ultimately shrinks the tax base. Governments then shrink into austerity mode, reinforcing the weakness in potential growth and accelerating the downward spiral.

We all know that breaking out of this vicious cycle requires more than incremental
adjustments; it demands radical, innovative change.

Who knows whether the Burnham administration will, or can, deliver the required policy changes within a timeframe, and with consequences, that are acceptable to an impatient, irascible electorate. Initial signs are encouraging but there is no guarantee that they will not end with a whimper. 

What follows is a humble economist's attempt to identify a structural reform agenda that addresses the macro changes that need to happen. It does not claim originality of policy proposals but, in exploring feasible outcomes through macro-finance models, reveals that we can dare to dream. 

## S.T.E.P. Forward

Enter a hypothetical **S.T.E.P.** platform (*Switch, Tech, Empower,
Purpose*) — a package of four targeted structural reforms designed for
short-term implementation (Years 1–3) as a foundation for a long-run growth revival without fiscal headaches:

1.  **S – Tax Switch for Working Families:** Rebalance the tax burden from income to wealth. This includes cutting employee National Insurance contributions, funded by equalising the tax treatment of unearned investment and capital income with earned income. Land taxes will also play a role in re-incentivising work and thus labour force participation.
2.  **T – Tech & AI Growth Guarantee:** Put AI, biotech and science at the heart of government, education and worker skillsets. Deploy co-investment (public/private, UK/European) vehicles via a National Wealth Fund and a British Business Bank. Take adavantage of homegrown expertise in AI, biotech, quantum computing and financial services to maximise employment opportinities, and above all, significantly raise productivity growth. 
3.  **E – Empowering Regions:** Fully activate and empower the revived National Economic Council. Joined-up thinking and action across regions, coupled with enhanced legitimacy through devolution, will raise productivity and boost both business and consumer confidence. 
4.  **P – Purpose-Driven Welfare (Health-to-Work Gateway):** Replace
    duplicate, adversarial assessment processes with a simplified, single
    Health-to-Work gateway. Substitute low-level cash top-ups with direct-access vouchers and
    immediate occupational therapy triage.

## The Self-Financing Virtuous Cycle

```{mermaid}
%%{init: {
  'theme': 'base',
  'themeVariables': {
    'fontSize': '32px',
    'fontFamily': 'Arial, sans-serif',
    'primaryColor': '#dceef8',
    'primaryTextColor': '#102a43',
    'primaryBorderColor': '#2980b9',
    'lineColor': '#2980b9',
    'edgeLabelBackground': '#ffffff',
    'clusterBkg': 'transparent',
    'clusterBorder': '#2980b9'
  },
  'flowchart': {
    'htmlLabels': true,
    'nodeSpacing': 30,
    'rankSpacing': 40
  }
}}%%

flowchart LR
    subgraph STEP["THE S.T.E.P. PILLARS"]
        direction TB
        S["SWITCH<br/>Tax Rebalancing"]
        T["TECH<br/>(GenAI TFP)"]
        E["EMPOWER<br/>Devolution"]
        P["PURPOSE<br/>Welfare Reform"]
    end

    S --> Hope
    T --> Hope
    E --> Hope
    P --> Hope

    subgraph CYCLE[" "]
        direction TB
        Hope["Boost Sentiment<br/>& Hope"]
        Growth["Accelerate Potential<br/>Real GDP Growth"]
        TaxBase["Expand Tax Base<br/>& Structural Savings"]
        Vigilantes["Ward Off Bond Vigilantes<br/>Real Gilt Yields fall"]
        Decay["Sustainable Debt Decay<br/>GDP Growth exceeds financing rate"]

        Hope --> Growth
        Growth --> TaxBase
        TaxBase --> Vigilantes
        Vigilantes --> Decay
    end

    classDef processBox fill:#dceef8,stroke:#2980b9,stroke-width:2px,color:#102a43;
    class S,T,E,P,Hope,Growth,TaxBase,Vigilantes,Decay processBox;

    style STEP fill:transparent,stroke:#2980b9,stroke-width:2px,color:#102a43
    style CYCLE fill:transparent,stroke:none
```


The core attraction of this S.T.E.P. platform lies in its **inbuilt
self-financing design** together with its capacity to break out the doom-loop and enter a reinforcing virtuous cycle. 

Crucially, as confidence returns, a positive feedback loop takes hold.
Improved sentiment crowds in private spending, labour force
participation stabilises via health-to-work pathways, and GenAI adoption
lifts baseline productivity. The strategy protects public finances in the short term and builds a solid foundation for future sustainability.

The resulting expansion of the real tax
base, combined with falling sovereign borrowing yields
and narrowing risk premia, wards off bond vigilantes and secures a sustainable, long-run fiscal
dividend.

------------------------------------------------------------------------

# The Macro Modelling Framework

To demonstrate that a positive outcome is feasible, we model the policy platform
in the context of standard growth and public finance models.

## Supply-Side Potential Growth

Aggregate output $Y_t$ is governed by a Cobb-Douglas production function
with labour-augmenting technological progress $A_t$:

$$Y_t = K_t^\alpha (A_t L_t)^{1-\alpha}$$

Taking logarithms and time derivatives yields the standard growth
accounting identity for potential real GDP growth ($g_t$):

$$g_t = \alpha g_{K,t} + (1-\alpha)(g_{A,t} + g_{L,t})$$

where:<br>
$\alpha = 0.40$ represents the profit share (capital share of
aggregate income), calibrated from Penn World Table (PWT) 11.0 national
income accounts<br><br>
$g_{A,t}$ is labour-augmenting Total Factor
Productivity (TFP) growth, which accelerates asymptotically from an
initial OBR baseline of 0.9% to a GenAI-boosted steady-state target of
**2.25%**<br><br>
$g_{L,t}$ is labour force growth, stabilised at
**0.4%–0.5%** via PIP health-to-work reforms<br><br>
$g_{K,t}$ is capital stock growth, which endogenously converges toward the Solow Balanced
Growth Path (BGP) equilibrium: $g_K^* = g_A + g_L = 2.65\%$.

## Capital Returns & Gilt Yields

The gross Marginal Product of Capital ($\text{MPK}$) and the net private
real return ($r_{\text{priv}}$) are derived from the aggregate
capital-output ratio ($K/Y$ \approx 4.8$, with this value based on PWT 11.0 estimates for the UK:

$$\text{MPK}_t = \alpha \left(\frac{Y_t}{K_t}\right) \approx 0.40 \times \left(\frac{1}{4.8}\right) \approx 8.3\%$$

$$r_{\text{priv},t} = \text{MPK}_t - \delta = 8.3\% - 5.0\% = 3.3\%$$

where $\delta = 5.0\%$ is the assumed fixed rate of capital depreciation.

Public sector real borrowing yields ($r_{\text{gov},t}$) differ from
private returns due to a market risk/convenience wedge ($\pi_{\text{risk},t}$):

$$r_{\text{priv},t} = r_{\text{gov},t} + \pi_{\text{risk},t}$$

Initial conditions for $r_{\text{gov}} = 3.1\%$ use the current current
10-year nominal gilt yield of \~5.1% less assumed inflation expectations of 2.0%.

$r_{\text{gov}}$ falls toward 1.0% as fiscal credibility is established with both domestic and international financial markets.

## Debt Sustainability Mechanics

The sovereign debt-to-GDP ratio ($b_t$) evolves dynamically:

$$b_t = (1 + r_{\text{gov},t} - g_t) b_{t-1} + pd_t$$

$$\Delta b_t = (r_{\text{gov},t} - g_t) b_{t-1} + pd_t$$

where $pd_t$ is the primary deficit ratio. 
The **Sustainability Gap** is defined as $g_t - r_{\text{gov},t}$.
Currently, Britain is in an unstable regime, signified by a negative gap. But, as the S.T.E.P program takes effect, the gap reverses into positive territory **$g > r_{\text{gov}}$. The economy then generates a positive growth dividend that naturally erodes the debt ratio, allowing structural primary surpluses ($pd < 0$) to accelerate debt reduction.

------------------------------------------------------------------------

## Simulation Results & Horizon Projections

The following simulation is calibrated using [Office for Budget Responsibility](https://obr.uk/frs/fiscal-risks-and-sustainability-july-2026/) (OBR) baseline figures for 2025/26 and estimates made in [Penn World Tables 11.0](https://www.rug.nl/ggdc/productivity/pwt/?lang=en)  

Initial conditions are as follows: $b_{2026} = 94.0\%$ of GDP, initial primary deficit $= 1.4\%$ of GDP, debt interest payments $= 2.9\%$ of GDP, GDP growth $= 1.5\%$yoy.

As expected, a successful and sustained S.T.E.P. implementation leads the UK toward a desirable long-run steady state - <b>a happy place</b>.

```{r}
library(tidyverse)
library(gt)
library(patchwork)

# ------------------------------------------------------------------------------
# 1. Base Model Calibration Parameters
# ------------------------------------------------------------------------------
alpha        <- 0.40   # Profit share / Capital share of income (PWT 11.0)
delta        <- 0.05   # Annual capital depreciation rate (5%)
pi_target    <- 0.02   # Expected inflation target (2%)
b_initial    <- 0.940  # OBR starting debt ratio at end-2026 (94.0%)
ky_init      <- 4.80   # Capital-Output Ratio (PWT 11.0 baseline)

macro_sim <- tibble(
  Period = c("2027", "2028", "2029", "2030s (Avg)", "Long Run (2040s+)"),
  tfp_growth      = c(0.90, 1.35, 1.70, 2.25, 2.25) / 100, # TFP growth
  labour_growth   = c(0.60, 0.45, 0.45, 0.50, 0.40) / 100, # Labour supply
  r_gov           = c(3.10, 2.20, 1.75, 1.25, 1.00) / 100, # Public real funding rate
  primary_deficit = c(1.40, 0.60, 0.00, -0.60, -1.20) / 100, # Primary deficit
  ky_ratio        = c(4.80, 4.80, 4.80, 4.85, 4.90)        # Capital-Output ratio
)

# ------------------------------------------------------------------------------
# 2. Dynamic Growth Accounting & Chained Debt Calculations
# ------------------------------------------------------------------------------
macro_sim <- macro_sim %>%
  mutate(eff_labour_growth = tfp_growth + labour_growth)

gK_vec <- numeric(nrow(macro_sim))
gK_vec[1] <- 0.80 * macro_sim$eff_labour_growth[1]
for(t in 2:nrow(macro_sim)) {
  adj_weight <- ifelse(t == 5, 1.0, 0.65 + (t * 0.08))
  gK_vec[t]  <- adj_weight * macro_sim$eff_labour_growth[t] + (1 - adj_weight) * gK_vec[t-1]
}

macro_sim <- macro_sim %>%
  mutate(
    capital_growth = gK_vec,
    gdp_growth = (alpha * capital_growth) + ((1 - alpha) * eff_labour_growth),
    g_minus_r = gdp_growth - r_gov,
    mpk_gross = alpha * (1 / ky_ratio),
    r_private = mpk_gross - delta,
    risk_premium = r_private - r_gov
  )

n_periods <- nrow(macro_sim)
b_start <- numeric(n_periods)
b_end   <- numeric(n_periods)
current_b <- b_initial

for (i in 1:n_periods) {
  b_start[i] <- current_b
  horizon_years <- ifelse(i == 4, 10, ifelse(i == 5, 10, 1))
  temp_b <- current_b
  for (yr in 1:horizon_years) {
    temp_b <- (1 + macro_sim$r_gov[i] - macro_sim$gdp_growth[i]) * temp_b + macro_sim$primary_deficit[i]
  }
  b_end[i] <- temp_b
  current_b <- temp_b 
}

macro_sim <- macro_sim %>%
  mutate(
    debt_start = b_start,
    debt_end = b_end,
    debt_interest_pay = r_gov * debt_start,
    headline_deficit = primary_deficit + debt_interest_pay
  )
```

```{r}
#| label: macro-dashboard
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Macroeconomic Growth Acceleration and Fiscal Sustainability Trajectory"
#| fig-alt: "Two panel chart showing Solow potential growth convergence and declining government debt dynamics."

library(tidyverse)
library(gt)
library(patchwork)

theme_uk_policy <- function() {
  theme_minimal(base_size = 14) +
    theme(
      plot.title = element_text(face = "bold", size = 14, color = "#1a1a1a"),
      plot.subtitle = element_text(size = 12, color = "#4a4a4a"),
      axis.title = element_text(face = "bold", size = 12),
      axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_blank(),
      # panel.grid.major.x = element_line(color = "#e5e5e5"),
      # panel.grid.major.y = element_line(color = "#e5e5e5"),
      plot.background = element_rect(fill = "white", color = NA)
    )
}

p1 <- ggplot(macro_sim, aes(x = Period, group = 1)) +
  geom_line(aes(y = gdp_growth * 100, color = "Real GDP Growth (g)"), linewidth = 2.0) +
  geom_point(aes(y = gdp_growth * 100, color = "Real GDP Growth (g)"), size = 5) +
  geom_line(aes(y = r_gov * 100, color = "Gov Real Funding Rate (r_gov)"), linewidth = 2.0, linetype = "dashed") +
  geom_point(aes(y = r_gov * 100, color = "Gov Real Funding Rate (r_gov)"), size = 5) +
  scale_color_manual(name = "", values = c("Real GDP Growth (g)" = "#002060", "Gov Real Funding Rate (r_gov)" = "#c0392b")) +
  labs(
    title = "A. Real Growth vs. Public Borrowing Yields",
    subtitle = "Positive sustainability gap (g > r_gov) established by 2029",
    x = NULL, y = "Percent (%)"
  ) +
  theme_uk_policy() +
  theme(legend.position = "bottom")

p2 <- ggplot(macro_sim, aes(x = Period, group = 1)) +
  geom_col(
    aes(
      y = debt_end * 100,
      fill = "Government Debt Ratio (% GDP)"
    ),
    width = 0.8,
    alpha = 0.9
  ) +
  scale_y_continuous(
    name = "% of GDP"
  ) +
  scale_fill_manual(
    name = NULL,
    values = c(
      "Government Debt Ratio (% GDP)" = "#8e44ad"
    )
  ) +
  labs(
    title = "B. Sovereign Debt Decay",
    subtitle = "Debt ratio unwinds cleanly from 97% to 55% of GDP",
    x = NULL
  ) +
  theme_uk_policy() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank()
  )

p1 | p2
```


```{r}
#| label: macro-gt-table
#| echo: false
#| warning: false
#| message: false

# ------------------------------------------------------------------------------
# 1. Construct Transposed Data Frame for Table Output
# ------------------------------------------------------------------------------
macro_transposed <- tibble(
  Category = c(
    rep("Growth & Supply Side %yoy", 4),
    rep("Real Interest Rates %", 3),
    rep("Government Finances % of GDP", 5)
  ),
  Variable = c(
    "Real GDP", "Labour Force", "Capital Stock", "Total Factor Productivity",
    "Private Sector Net Real Return", "Risk Premium", "Government Real Funding Rate",
    "Primary Deficit", "Debt Interest Costs", "Headline Deficit", "Debt Ratio, end-period", "Sustainability Gap"
  ),
  `2027` = c(
    macro_sim$gdp_growth[1], macro_sim$labour_growth[1], macro_sim$capital_growth[1], macro_sim$tfp_growth[1],
    macro_sim$r_private[1], macro_sim$risk_premium[1], macro_sim$r_gov[1],
    macro_sim$primary_deficit[1], macro_sim$debt_interest_pay[1], macro_sim$headline_deficit[1], macro_sim$debt_end[1], macro_sim$g_minus_r[1]
  ) * 100,
  `2028` = c(
    macro_sim$gdp_growth[2], macro_sim$labour_growth[2], macro_sim$capital_growth[2], macro_sim$tfp_growth[2],
    macro_sim$r_private[2], macro_sim$risk_premium[2], macro_sim$r_gov[2],
    macro_sim$primary_deficit[2], macro_sim$debt_interest_pay[2], macro_sim$headline_deficit[2], macro_sim$debt_end[2], macro_sim$g_minus_r[2]
  ) * 100,
  `2029` = c(
    macro_sim$gdp_growth[3], macro_sim$labour_growth[3], macro_sim$capital_growth[3], macro_sim$tfp_growth[3],
    macro_sim$r_private[3], macro_sim$risk_premium[3], macro_sim$r_gov[3],
    macro_sim$primary_deficit[3], macro_sim$debt_interest_pay[3], macro_sim$headline_deficit[3], macro_sim$debt_end[3], macro_sim$g_minus_r[3]
  ) * 100,
  `2030s (Avg)` = c(
    macro_sim$gdp_growth[4], macro_sim$labour_growth[4], macro_sim$capital_growth[4], macro_sim$tfp_growth[4],
    macro_sim$r_private[4], macro_sim$risk_premium[4], macro_sim$r_gov[4],
    macro_sim$primary_deficit[4], macro_sim$debt_interest_pay[4], macro_sim$headline_deficit[4], macro_sim$debt_end[4], macro_sim$g_minus_r[4]
  ) * 100,
  `Long Run (2040s+)` = c(
    macro_sim$gdp_growth[5], macro_sim$labour_growth[5], macro_sim$capital_growth[5], macro_sim$tfp_growth[5],
    macro_sim$r_private[5], macro_sim$risk_premium[5], macro_sim$r_gov[5],
    macro_sim$primary_deficit[5], macro_sim$debt_interest_pay[5], macro_sim$headline_deficit[5], macro_sim$debt_end[5], macro_sim$g_minus_r[5]
  ) * 100
)

# ------------------------------------------------------------------------------
# 2. Construct Dynamic Footnote Text (With LaTeX Escaped % Symbols)
# ------------------------------------------------------------------------------
dynamic_source_note <- sprintf(
  "**Notes & Model Specifications:**<br><br>
   1. **Macro Supply:** Production technology assumes a Cobb-Douglas format with profit share $\\alpha = %.0f\\%%$, capital depreciation $\\delta = %.1f\\%%$, and $K/Y \\approx %.1f$. Gross $\\text{MPK} = \\alpha (Y/K) = %.1f\\%%$<br><br>
   2. **Debt & Inflation:** Inflation expectations anchored at $\\pi = %.1f\\%%$; corresponding nominal 10-year funding rate is $i = r_{\\text{gov}} + \\pi = %.1f\\%%$ in 2027<br><br>
   3. **Sustainability Gap:** Real GDP Growth less Government Real Funding Rate ($g - r_{\\text{gov}}$). Positive indicates dynamic stability ($g > r_{\\text{gov}}$)<br><br>
   4. **Solow Balanced Growth Path:** Final column represents the model's steady state<br> $g_K^* = g_A + g_L = %.2f\\%%$, yielding $g^* = %.2f\\%%$",

  alpha * 100,                                        
  delta * 100,                                        
  ky_init,                                            
  macro_sim$mpk_gross[1] * 100,                       
  pi_target * 100,                                    
  (macro_sim$r_gov[1] + pi_target) * 100,             
  (macro_sim$tfp_growth[5] +
     macro_sim$labour_growth[5]) * 100,               
  macro_sim$gdp_growth[5] * 100                       
)

# ------------------------------------------------------------------------------
# 3. Render Table directly to Quarto HTML Pipeline
# ------------------------------------------------------------------------------
gt_a4_table <- macro_transposed %>%
  gt(groupname_col = "Category") %>%
  tab_header(
    title = md("**Illustrative Macro-Fiscal Simulation**"),
    subtitle = md("Initialised using OBR and Penn World Tables 11.0 estimates")) %>%
  cols_label(Variable = "") %>%
  
  # --- Column Width Fine-Tuning ---
  # Adjust px or % values below to balance layout across desktop & mobile
  cols_width(
    Variable ~ px(260),
    `2027` ~ px(85),
    `2028` ~ px(85),
    `2029` ~ px(85),
    `2030s (Avg)` ~ px(110),
    `Long Run (2040s+)` ~ px(140)
  ) %>%
  
  # --- Number Formatting (With % suffix) ---
  fmt_number(
    columns = -c(Category, Variable), 
    decimals = 1,
    pattern = "{x}%"  # Appends % directly to table cell values (e.g. 1.4%)
  ) %>%
  
  # --- Header & Group Styles ---
  tab_style(
    style = cell_text(weight = "bold"), 
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#111827"), cell_text(weight = "bold", color = "#60a5fa")),
    locations = cells_row_groups(groups = "Growth & Supply Side %yoy")
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#111827"), cell_text(weight = "bold", color = "#f59e0b")),
    locations = cells_row_groups(groups = "Real Interest Rates %")
  ) %>%
  tab_style(
    style = list(cell_fill(color = "#111827"), cell_text(weight = "bold", color = "#34d399")),
    locations = cells_row_groups(groups = "Government Finances % of GDP")
  ) %>%
  tab_source_note(source_note = md(dynamic_source_note)) %>%
  tab_style(
    style = cell_text(
    size = px(16),
    color = "#111827"),
    locations = cells_source_notes()) %>%
  opt_css(css = ".gt_sourcenote {line-height: 1.2;}")

# Call the object directly for Quarto execution
gt_a4_table
```

# Take That
Britain could limp along towards the next election. But who wants that and the dangers it poses to social and economic life. There is clearly appetite for radical change and the S.T.E.P framework offers ideas. For those worried about whether it all adds up, the macro-finance simulations show that it can - in terms of combining strong growtyh, better living standards, stable inflation and sustainable public finances. Whether the current administarion is      

::: {.bibliography}

<p class="bibliography-title">References</p>

1. Resolution Foundation (2026),  
   [*See it. Say it. Sort it: How to reset Britain’s economic policy*](https://www.resolutionfoundation.org/).

2. IPPR (2026),  
   [*Turning energy support into investment leverage*](https://www.ippr.org/).

3. Joseph Rowntree Foundation (2026),  
   [*Tackle economic failures at root to ease pressure on Universal Credit*](https://www.jrf.org.uk/).
   
:::
